---
title: "CSC 587 HW 1"
author: "Daniel R. Getty"
date: "`r Sys.Date()`"
output: pdf_document
---
## Set Environment
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE)
# directory
dir <- 'G:\\My Drive\\H Drive\\Course Work\\CERG-Data Science\\CSC_587_Advanced_Data_Mining\\HW\\HW1_DataMining_with_R'
# Set the working directory.
setwd(dir)
# Print the working directory.
getwd()
# load ggplot2 package
library(ggplot2)
```

## Homework 1
1.a: Use read.delim function to read Su_raw_matrix.txt into a variable called 
su.(Notice that su has become a data frame now) Load in the data set from disk.
```{r 1.a }
data_file <- file.path('Su_raw_matrix.txt')
# Build data frame from the data set.
su <- read.delim(data_file, header = TRUE, sep = '\t')
# Print the data frame.
su
```
1.b: Use mean and sd functions to find mean and standard deviation of 
```{r 1.b Starndard Deviation}
sd(su$Liver_2.CEL)
```

```{r 1.b Mean}
mean(su$Liver_2.CEL)
```

1.c: Use colMeans and colSums functions to get the average and total values of 
each column.

Column Means
```{r 1.c Coumn Means}
colMeans(su[, -1])  # Exclude the first column
```
Column Sums
```{r 1.c Column Sums}
colSums(su[, -1])  # Exclude the first Column
```

2. Use rnorm(n, mean = 0, sd = 1) function to generate 10000 numbers for the following(mean, sigma) pairs and plot histogram for each, meaning you need to 
change the function parameter accordingly. Then comment on how these histograms 
are different from each other and state the reason. Please save your figures as 
image from RStudio. (Hint: to see the difference in plots you may need to set 
the xlim parameter in plot function to c(-5,5))

2.a: mean = 0, sigma = 0.2
```{r 2.a: mean = 0, sigma = 0.2}
hist(rnorm(10000, mean = 0, sd = 0.2),xlim = c(-5,5))

```

2.b: mean = 0, sigma = 0.5
```{r 2.b: mean = 0, sigma = 0.5} 
hist(rnorm(10000, mean = 0, sd = 0.5),xlim = c(-5,5))
```
The difference between the two histograms is the standard deviation. The first
being 0.2 and the second 0.5. "Standard Deviation (sigma,or σ) is a measure of 
how dispersed the data is in relation to the mean. Low, or small, standard 
deviation indicates data are clustered tightly around the mean, and high, or 
large, standard deviation indicates data are more spread out. A standard 
deviation close to zero indicates that data points are very close to the mean, whereas a larger standard deviation indicates data points are spread further 
away from the mean." (https://www.nlm.nih.gov/oet/ed/stats/02-900.html)

3. Perform the steps below with ”dat” dataframe which is just a sample data for 
you to observe how each plot function ( 3b through 3e ) works. Notice that you 
need to have ggplot2 library installed on your system. Please refer slides how
to install and import a library. Installation is done only once, but you need to import the library every time you need it by saying library(ggplot2). Then run 
the following commands for questions from 3a through 3e and observe how the 
plots are generated first. 

3.a: 
```{r 3.a}
dat <- data.frame(cond = factor(rep(c("A","B"), each=200)), rating = c(rnorm(200),rnorm(200, mean=.8)))
```

3.b: Overlaid histograms
```{r 3.b}
ggplot(dat, aes(x=rating, fill=cond)) + geom_histogram(binwidth=.5, alpha=.5, position="identity")
```
3.c: Interleaved Histograms
```{r 3.c}
ggplot(dat, aes(x=rating, fill=cond)) + geom_histogram(binwidth=.5, position="dodge")
```

3.d: Density plots
```{r 3.d}
ggplot(dat, aes(x=rating, colour=cond)) + geom_density()

```

3.e: Density plots with semitransparent fill
```{r 3.e}
ggplot(dat, aes(x=rating, fill=cond)) + geom_density(alpha=.3)
```

3.f: Read ”diabetes_train.csv” into a variable called diabetes and apply the same functions 3b through 3e for the mass attribute of diabetes and save the images. (Hint: instead of cond above, use the class attribute to color your groups. When 
you have fill option, your plots should show same type of chart for both groups 
in different colors on the same figure. Keep in mind that diabetes and dat are 
both DataFrames)

Load Diabetes Data
```{r 3.f}
diabetes_fp <- file.path('diabetes.csv')
# Build data frame from the data set.
diabetes <- read.csv(diabetes_fp, header = TRUE, sep = ',')

dat <- data.frame(cond = factor(rep(c("A","B"), each=200)), rating = c(rnorm(200),rnorm(200, mean=.8)))

pregmean <- mean(diabetes$Pregnancies)
pregsd <- sd(diabetes$Pregnancies)
rnormpreg=rnorm(diabetes$Pregnancies, mean=pregmean, sd=pregsd)

dbpreg <- data.frame(preg = diabetes$Pregnancies, rnormpreg, outcome=diabetes$Outcome)

dbpreg
```
Overlaid histograms
```{r 3.f}
# ggplot(dat, aes(x=rating, fill=cond)) + geom_histogram(binwidth=.5, alpha=.5, position="identity")

ggplot(dbpreg, aes(x=rnormpreg, fill=outcome)) + geom_histogram(binwidth=.5, alpha=.5, position="identity")
```